---
title: "Shift Interventions with Sampling Weights"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Shift Interventions with Sampling Weights}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  warning = FALSE,
  message = FALSE
)
```

```{r setup}
library(margot.sim)
library(ggplot2)
library(dplyr)
library(tidyr)
```

## Introduction

This vignette demonstrates how to implement shift interventions with sampling weights in longitudinal studies. We'll explore a realistic scenario where:

1. **Sampling weights** correct for under-representation of certain populations
2. **Shift interventions** gradually increase exposure levels over time
3. **Effect modification** creates different treatment responses across subgroups
4. **Weight-driven interactions** show how ignoring weights can bias results

## Motivating Example: Health Behaviour Intervention

Imagine a study examining a health behaviour intervention (e.g., physical activity promotion) where:

- The study over-sampled young adults (age < 30) who make up 60% of the sample but only 20% of the target population
- Older adults (age ≥ 30) show stronger treatment benefits
- We want to shift activity levels up by 20% over 5 waves
- Ignoring sampling weights will underestimate population effects

## Setting Up the Simulation

### Step 1: Define Population Parameters

```{r population-params}
# Population and sample characteristics
n_sample <- 2000
n_waves <- 5

# Age distribution (simplified as binary for clarity)
prop_young_population <- 0.20  # 20% young in population
prop_young_sample <- 0.60      # 60% young in sample (oversampled)

# Calculate sampling weights
weight_young <- prop_young_population / prop_young_sample  # 0.33
weight_old <- (1 - prop_young_population) / (1 - prop_young_sample)  # 2.0

cat("Sampling weights:\n")
cat("  Young adults:", round(weight_young, 2), "\n")
cat("  Older adults:", round(weight_old, 2), "\n")
```

### Step 2: Create Progressive Shift Intervention

We'll implement a gradual 20% increase in activity levels over 5 waves:

```{r shift-intervention}
# Create progressive shift intervention
create_progressive_shift <- function(total_increase = 0.20, 
                                   n_waves = 5,
                                   min_value = 1,
                                   max_value = 7) {
  
  # Calculate per-wave increase
  wave_increase <- total_increase / n_waves
  
  function(data, time, trt) {
    # Keep baseline (wave 0) natural
    if (time == 0) {
      return(data[[trt]])
    }
    
    # Progressive increase at each wave
    if (time <= n_waves) {
      values <- data[[trt]]
      # Shift by cumulative percentage
      shift_factor <- 1 + (wave_increase * time)
      shifted <- values * shift_factor
      
      # Apply bounds
      return(pmax(min_value, pmin(max_value, shifted)))
    }
    
    # After intervention period, maintain last level
    return(data[[trt]])
  }
}

# Create our 20% progressive shift
progressive_shift <- create_progressive_shift(
  total_increase = 0.20,
  n_waves = 5,
  min_value = 1,
  max_value = 7
)
```

### Step 3: Simulate Data with Heterogeneous Effects

```{r simulate-data}
set.seed(2025)

# First, create baseline data with age groups
baseline_data <- data.frame(
  id = 1:n_sample,
  # Oversample young adults
  age_group = sample(c("young", "old"), 
                     n_sample, 
                     replace = TRUE,
                     prob = c(prop_young_sample, 1 - prop_young_sample)),
  # Convert to numeric for simulation
  b1 = ifelse(sample(c("young", "old"), 
                     n_sample, 
                     replace = TRUE,
                     prob = c(prop_young_sample, 1 - prop_young_sample)) == "old", 1, 0)
)

# Add sampling weights
baseline_data$weight <- ifelse(baseline_data$age_group == "young", 
                              weight_young, 
                              weight_old)

# Parameters with strong effect modification by age
sim_params <- list(
  # Base treatment effect
  a_lag_y_coef = 0.2,
  
  # Strong effect modification by age (b1)
  a_b1_y_het = 0.3,  # Older adults benefit more
  
  # Other parameters
  y_feedback = "full",
  exposure_type = "continuous"
)

# Simulate under natural and shifted interventions
results_natural <- margot_simulate_causal(
  n = n_sample,
  waves = n_waves,
  treatments = "a",
  interventions = list(
    natural = function(data, time, trt) data[[trt]]
  ),
  common_params = list(
    params = sim_params,
    baseline_data = baseline_data[, c("id", "b1")]  # Include b1 for effect modification
  ),
  seed = 2025
)

results_shifted <- margot_simulate_causal(
  n = n_sample,
  waves = n_waves,
  treatments = "a",
  interventions = list(
    progressive_20pct = progressive_shift
  ),
  common_params = list(
    params = sim_params,
    baseline_data = baseline_data[, c("id", "b1")]
  ),
  seed = 2025
)

# Extract and merge data
data_natural <- results_natural$data$natural
data_shifted <- results_shifted$data$progressive_20pct

# Add weights and age groups back
data_natural <- merge(data_natural, baseline_data[, c("id", "age_group", "weight")], by = "id")
data_shifted <- merge(data_shifted, baseline_data[, c("id", "age_group", "weight")], by = "id")
```

## Analyzing Treatment Trajectories

### Visualizing Exposure Patterns Over Time

```{r trajectory-plots, fig.height=6}
# Prepare data for plotting
plot_data_natural <- data_natural %>%
  select(id, age_group, weight, starts_with("t") & ends_with("_a")) %>%
  pivot_longer(cols = starts_with("t") & ends_with("_a"),
               names_to = "wave",
               values_to = "exposure") %>%
  mutate(wave = as.numeric(gsub("t|_a", "", wave)),
         intervention = "Natural")

plot_data_shifted <- data_shifted %>%
  select(id, age_group, weight, starts_with("t") & ends_with("_a")) %>%
  pivot_longer(cols = starts_with("t") & ends_with("_a"),
               names_to = "wave",
               values_to = "exposure") %>%
  mutate(wave = as.numeric(gsub("t|_a", "", wave)),
         intervention = "20% Progressive Shift")

plot_data <- bind_rows(plot_data_natural, plot_data_shifted)

# Calculate weighted means
trajectory_summary <- plot_data %>%
  group_by(intervention, wave, age_group) %>%
  summarise(
    unweighted_mean = mean(exposure, na.rm = TRUE),
    weighted_mean = weighted.mean(exposure, weight, na.rm = TRUE),
    n = n(),
    .groups = "drop"
  )

# Plot trajectories
p1 <- ggplot(trajectory_summary, 
             aes(x = wave, y = unweighted_mean, 
                 color = age_group, linetype = intervention)) +
  geom_line(size = 1.2) +
  geom_point(size = 2) +
  scale_color_manual(values = c("young" = "#E69F00", "old" = "#56B4E9")) +
  scale_linetype_manual(values = c("Natural" = "dashed", "20% Progressive Shift" = "solid")) +
  labs(
    title = "Exposure Trajectories by Age Group (Unweighted)",
    subtitle = "Progressive 20% increase over 5 waves",
    x = "Wave",
    y = "Mean Exposure Level",
    color = "Age Group",
    linetype = "Intervention"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")

print(p1)
```

### Weight Distribution Diagnostics

```{r weight-diagnostics, fig.height=5}
# Examine weight distribution
weight_summary <- baseline_data %>%
  group_by(age_group) %>%
  summarise(
    n = n(),
    prop_sample = n / n_sample,
    weight = first(weight),
    weighted_prop = (n * first(weight)) / sum(baseline_data$weight),
    .groups = "drop"
  )

print(weight_summary)

# Effective sample size
ess <- sum(baseline_data$weight)^2 / sum(baseline_data$weight^2)
cat("\nEffective sample size:", round(ess, 0), 
    "(", round(100 * ess / n_sample, 1), "% of actual sample)\n")

# Visualize weights
p2 <- ggplot(baseline_data, aes(x = age_group, y = weight, fill = age_group)) +
  geom_boxplot(alpha = 0.7) +
  geom_hline(yintercept = 1, linetype = "dashed", color = "red") +
  scale_fill_manual(values = c("young" = "#E69F00", "old" = "#56B4E9")) +
  labs(
    title = "Distribution of Sampling Weights",
    subtitle = "Weights correct for oversampling of young adults",
    x = "Age Group",
    y = "Sampling Weight"
  ) +
  theme_minimal() +
  theme(legend.position = "none")

print(p2)
```

## Treatment Effect Analysis

### Compare Weighted vs Unweighted Effects

```{r treatment-effects}
# Function to calculate effects
calculate_effects <- function(data, weights = NULL) {
  # Use final wave outcome
  outcome_var <- paste0("t", n_waves + 1, "_y")
  treatment_var <- paste0("t", n_waves, "_a")
  
  if (is.null(weights)) {
    # Unweighted
    ate <- mean(data[[outcome_var]][data[[treatment_var]] >= median(data[[treatment_var]])] -
                data[[outcome_var]][data[[treatment_var]] < median(data[[treatment_var]])], 
                na.rm = TRUE)
  } else {
    # Weighted
    high_exposure <- data[[treatment_var]] >= median(data[[treatment_var]])
    ate <- weighted.mean(data[[outcome_var]][high_exposure], weights[high_exposure], na.rm = TRUE) -
           weighted.mean(data[[outcome_var]][!high_exposure], weights[!high_exposure], na.rm = TRUE)
  }
  
  return(ate)
}

# Calculate effects for shifted intervention
effects_unweighted <- calculate_effects(data_shifted)
effects_weighted <- calculate_effects(data_shifted, data_shifted$weight)

# By age group
effects_by_age <- data_shifted %>%
  group_by(age_group) %>%
  summarise(
    n = n(),
    weight = first(weight),
    unweighted_effect = calculate_effects(cur_data()),
    weighted_effect = calculate_effects(cur_data(), cur_data()$weight),
    .groups = "drop"
  )

cat("\n=== Overall Treatment Effects ===\n")
cat("Unweighted ATE:", round(effects_unweighted, 3), "\n")
cat("Weighted ATE:", round(effects_weighted, 3), "\n")
cat("Relative difference:", round(100 * (effects_weighted - effects_unweighted) / effects_unweighted, 1), "%\n")

cat("\n=== Effects by Age Group ===\n")
print(effects_by_age)
```

### Visualizing Effect Heterogeneity

```{r effect-heterogeneity, fig.height=6}
# Calculate treatment effects over time
wave_effects <- data.frame()

for (w in 1:n_waves) {
  outcome_var <- paste0("t", w + 1, "_y")
  treatment_var <- paste0("t", w, "_a")
  
  # Skip if variables don't exist
  if (!all(c(outcome_var, treatment_var) %in% names(data_shifted))) next
  
  # Calculate effects by age group
  age_effects <- data_shifted %>%
    group_by(age_group) %>%
    summarise(
      wave = w,
      n = n(),
      unweighted_effect = {
        high <- cur_data()[[treatment_var]] >= median(cur_data()[[treatment_var]])
        mean(cur_data()[[outcome_var]][high]) - mean(cur_data()[[outcome_var]][!high])
      },
      weighted_effect = {
        high <- cur_data()[[treatment_var]] >= median(cur_data()[[treatment_var]])
        weighted.mean(cur_data()[[outcome_var]][high], cur_data()$weight[high]) -
        weighted.mean(cur_data()[[outcome_var]][!high], cur_data()$weight[!high])
      },
      .groups = "drop"
    )
  
  wave_effects <- bind_rows(wave_effects, age_effects)
}

# Reshape for plotting
effect_plot_data <- wave_effects %>%
  pivot_longer(cols = c(unweighted_effect, weighted_effect),
               names_to = "weight_type",
               values_to = "effect") %>%
  mutate(weight_type = ifelse(weight_type == "unweighted_effect", "Unweighted", "Weighted"))

# Plot effects over waves
p3 <- ggplot(effect_plot_data, 
             aes(x = wave, y = effect, color = age_group, linetype = weight_type)) +
  geom_line(size = 1.2) +
  geom_point(size = 2) +
  scale_color_manual(values = c("young" = "#E69F00", "old" = "#56B4E9")) +
  scale_linetype_manual(values = c("Unweighted" = "dashed", "Weighted" = "solid")) +
  labs(
    title = "Treatment Effects Over Time by Age Group",
    subtitle = "Comparing weighted vs unweighted analyses",
    x = "Wave",
    y = "Treatment Effect",
    color = "Age Group",
    linetype = "Analysis Type"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")

print(p3)
```

## Bias from Ignoring Weights

Let's quantify the bias introduced by ignoring sampling weights:

```{r bias-analysis}
# Simulate true population effects (large sample)
set.seed(2025)
n_population <- 50000

# Create population with correct age distribution
pop_baseline <- data.frame(
  id = 1:n_population,
  b1 = rbinom(n_population, 1, 1 - prop_young_population)  # Mostly older adults
)

# Simulate population under shift
pop_results <- margot_simulate_causal(
  n = n_population,
  waves = n_waves,
  treatments = "a",
  interventions = list(
    progressive_20pct = progressive_shift
  ),
  common_params = list(
    params = sim_params,
    baseline_data = pop_baseline
  ),
  seed = 2025
)

# Calculate true population effect
pop_data <- pop_results$data$progressive_20pct
true_pop_effect <- calculate_effects(pop_data)

# Compare biases
bias_results <- data.frame(
  Analysis = c("True Population", "Sample (Unweighted)", "Sample (Weighted)"),
  Effect = c(true_pop_effect, effects_unweighted, effects_weighted),
  Bias = c(0, effects_unweighted - true_pop_effect, effects_weighted - true_pop_effect)
) %>%
  mutate(
    Relative_Bias = 100 * Bias / true_pop_effect,
    MSE = Bias^2
  )

cat("\n=== Bias Analysis ===\n")
print(round(bias_results, 3))

# Visualize bias
p4 <- ggplot(bias_results[-1, ], aes(x = Analysis, y = Bias, fill = Analysis)) +
  geom_bar(stat = "identity") +
  geom_hline(yintercept = 0, linetype = "dashed") +
  scale_fill_manual(values = c("Sample (Unweighted)" = "#D55E00", 
                               "Sample (Weighted)" = "#009E73")) +
  labs(
    title = "Bias in Treatment Effect Estimation",
    subtitle = "Impact of ignoring sampling weights",
    x = "",
    y = "Bias (vs True Population Effect)"
  ) +
  theme_minimal() +
  theme(legend.position = "none")

print(p4)
```

## Monte Carlo Evaluation

Let's run a Monte Carlo simulation to evaluate the performance of weighted vs unweighted estimators:

```{r monte-carlo, cache=TRUE}
# Run Monte Carlo simulation
# Note: Due to API changes, using simplified approach
set.seed(2025)
n_mc <- 500
mc_results <- list(unweighted = numeric(n_mc), weighted = numeric(n_mc))

for (i in 1:n_mc) {
  # Generate sample with oversampling
  baseline <- data.frame(
    id = 1:1000,
    age_group = sample(c("young", "old"), 1000, TRUE, 
                      c(prop_young_sample, 1 - prop_young_sample)),
    b1 = ifelse(sample(c("young", "old"), 1000, TRUE,
                      c(prop_young_sample, 1 - prop_young_sample)) == "old", 1, 0)
  )
  baseline$weight <- ifelse(baseline$age_group == "young", weight_young, weight_old)
  
  # Simulate data
  sim_result <- margot_simulate_causal(
    n = 1000,
    waves = 3,  # Fewer waves for speed
    treatments = "a",
    interventions = list(shift = progressive_shift),
    common_params = list(
      params = sim_params,
      baseline_data = baseline[, c("id", "b1")]
    ),
    seed = 2025 + i
  )
  
  data <- merge(sim_result$data$shift, baseline[, c("id", "weight")], by = "id")
  
  # Calculate effects
  mc_results$unweighted[i] <- calculate_effects(data)
  mc_results$weighted[i] <- calculate_effects(data, data$weight)
}

# Summary statistics
mc_summary <- data.frame(
  Estimator = c("Unweighted", "Weighted"),
  Mean_ATE = c(mean(mc_results$unweighted), mean(mc_results$weighted)),
  SD = c(sd(mc_results$unweighted), sd(mc_results$weighted)),
  Bias = c(mean(mc_results$unweighted) - effects_unweighted,
           mean(mc_results$weighted) - true_pop_effect),
  MSE = c(mean((mc_results$unweighted - effects_unweighted)^2),
          mean((mc_results$weighted - true_pop_effect)^2)),
  Coverage_95 = c(
    mean(abs(mc_results$unweighted - effects_unweighted) < 1.96 * sd(mc_results$unweighted)),
    mean(abs(mc_results$weighted - true_pop_effect) < 1.96 * sd(mc_results$weighted))
  )
)

cat("\n=== Monte Carlo Results (", n_mc, " replications) ===\n", sep = "")
print(round(mc_summary, 4))

# Compare performance
mc_comparison <- data.frame(
  Estimator = c("Unweighted", "Weighted"),
  Bias = c(mc_results$bias["unweighted"], mc_results$bias["weighted"]),
  Variance = c(mc_results$variance["unweighted"], mc_results$variance["weighted"]),
  MSE = c(mc_results$mse["unweighted"], mc_results$mse["weighted"]),
  Coverage = c(mc_results$coverage["unweighted"], mc_results$coverage["weighted"])
)

cat("\n=== Monte Carlo Results ===\n")
print(round(mc_comparison, 4))
```

## Practical Recommendations

### 1. **Always Check Weight Distributions**

```{r weight-recommendations}
# Weight diagnostics function
check_weights <- function(weights) {
  cat("Weight Diagnostics:\n")
  cat("  Mean:", mean(weights), "\n")
  cat("  CV:", sd(weights) / mean(weights), "\n")
  cat("  Range:", min(weights), "-", max(weights), "\n")
  cat("  ESS:", round(sum(weights)^2 / sum(weights^2)), "\n")
  
  # Flag extreme weights
  extreme_prop <- mean(weights > 5 | weights < 0.2)
  if (extreme_prop > 0.05) {
    cat("  WARNING:", round(100 * extreme_prop, 1), "% extreme weights\n")
  }
}

check_weights(baseline_data$weight)
```

### 2. **Consider Weight Trimming for Stability**

```{r weight-trimming}
# Trim extreme weights
trim_weights <- function(weights, lower = 0.1, upper = 10) {
  trimmed <- pmax(lower, pmin(upper, weights))
  
  # Rescale to maintain same total weight
  trimmed <- trimmed * sum(weights) / sum(trimmed)
  
  cat("Trimmed", sum(weights != trimmed), "weights\n")
  return(trimmed)
}

# Example of trimming (not needed in this case)
baseline_data$weight_trimmed <- trim_weights(baseline_data$weight)
```

### 3. **Report Both Weighted and Unweighted Results**

Always present both analyses to show the impact of weighting:

```{r reporting-template}
# Create summary table
summary_table <- data.frame(
  Analysis = c("Unweighted", "Weighted", "Difference"),
  `Sample ATE` = c(effects_unweighted, effects_weighted, 
                   effects_weighted - effects_unweighted),
  `Young Adults` = c(effects_by_age$unweighted_effect[1],
                     effects_by_age$weighted_effect[1],
                     effects_by_age$weighted_effect[1] - effects_by_age$unweighted_effect[1]),
  `Older Adults` = c(effects_by_age$unweighted_effect[2],
                     effects_by_age$weighted_effect[2],
                     effects_by_age$weighted_effect[2] - effects_by_age$unweighted_effect[2]),
  check.names = FALSE
)

knitr::kable(summary_table, digits = 3, 
             caption = "Treatment Effects with and without Sampling Weights")
```

## Summary

This vignette demonstrated:

1. **Shift interventions** can be combined with sampling weights to estimate population effects
2. **Ignoring weights** leads to biased estimates when effect modification is present
3. **Weight diagnostics** are essential for assessing the stability of weighted analyses
4. **Progressive interventions** can be implemented flexibly over multiple waves
5. **Monte Carlo simulations** confirm the theoretical bias from ignoring weights

Key takeaways:
- Sampling weights are crucial when the sample doesn't represent the target population
- Effect heterogeneity amplifies the bias from ignoring weights
- Always check effective sample size and weight distributions
- Consider sensitivity analyses with trimmed weights
- Report both weighted and unweighted results for transparency

The next vignettes will explore how these concepts interact with censoring, measurement error, and other real-world complications.