---
title: "Practical Workflow: Complete Analysis Pipeline"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Practical Workflow: Complete Analysis Pipeline}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 6,
  warning = FALSE,
  message = FALSE
)
```

```{r setup}
library(margot.sim)
library(ggplot2)
library(dplyr)
library(tidyr)
library(knitr)
```

## Introduction

This vignette provides a complete, practical workflow for using margot.sim in a realistic research scenario. We'll walk through:

1. **Study design** and parameter specification
2. **Data generation** with realistic features
3. **Shadow application** for observational challenges
4. **Analysis pipeline** with multiple estimators
5. **Sensitivity analyses** and diagnostics
6. **Reporting** and interpretation

## Scenario: Evaluating a Workplace Mental Health Program

We're evaluating a workplace wellness program designed to improve mental health through:
- Mindfulness training and stress reduction techniques
- Flexible work arrangements for those who need them
- Gradual implementation over 12 months (4 quarters)

Key challenges:
- Some employees more likely to participate (selection)
- Mental health measures subject to reporting bias (measurement error)
- Higher-stress employees more likely to leave (censoring)
- Effects vary by baseline stress and job type

## Step 1: Study Design and Planning

```{r study-design}
# Study parameters
n_employees <- 3000
n_waves <- 4  # Quarterly measurements
baseline_date <- "2024-01-01"

# Create study design document
study_design <- list(
  title = "Workplace Mental Health Intervention Evaluation",
  population = "All employees at participating companies",
  sample_size = n_employees,
  waves = n_waves,
  measurement_schedule = "Quarterly (0, 3, 6, 9, 12 months)",
  primary_outcome = "Mental health score (0-10 scale)",
  treatment = "Comprehensive wellness program",
  key_confounders = c(
    "Baseline stress level (b1)",
    "Job type - client-facing vs internal (b2)", 
    "Years at company (b3)"
  )
)

cat("=== Study Design Summary ===\n")
for (element in names(study_design)) {
  cat(element, ":", study_design[[element]], "\n")
}

# Power calculation (simplified)
expected_effect <- 0.3  # Expected improvement
expected_sd <- 1.2      # Expected standard deviation
power_n <- (4 * expected_sd^2 * (qnorm(0.975) + qnorm(0.80))^2) / expected_effect^2
cat("\nRequired sample size for 80% power:", round(power_n), "\n")
cat("Planned sample size:", n_employees, "✓\n")
```

## Step 2: Generate Realistic Base Data

```{r generate-data}
set.seed(2025)

# Create realistic baseline characteristics
generate_workforce_data <- function(n) {
  data.frame(
    id = 1:n,
    # Company (for clustering)
    company = sample(LETTERS[1:5], n, replace = TRUE,
                    prob = c(0.3, 0.25, 0.2, 0.15, 0.1)),
    # Job characteristics
    job_type = sample(c("Client-facing", "Internal"), n, TRUE, c(0.4, 0.6)),
    years_employed = rexp(n, rate = 0.2),  # Average 5 years
    department = sample(c("Sales", "Tech", "Admin", "Management"), n, TRUE)
  ) %>%
  mutate(
    # Convert to simulation variables
    b1 = rnorm(n),  # Baseline stress (standardized)
    b2 = ifelse(job_type == "Client-facing", 1, -1) + rnorm(n, 0, 0.3),
    b3 = scale(years_employed)[,1],  # Standardized years
    # Add clustering by company
    company_effect = rep(rnorm(5, 0, 0.3)[match(company, LETTERS[1:5])], length.out = n),
    b1 = b1 + company_effect  # Companies have different stress cultures
  )
}

workforce_data <- generate_workforce_data(n_employees)

# Examine baseline characteristics
baseline_summary <- workforce_data %>%
  group_by(company) %>%
  summarise(
    n = n(),
    pct_client_facing = mean(job_type == "Client-facing") * 100,
    mean_years = mean(years_employed),
    mean_stress = mean(b1),
    .groups = "drop"
  )

cat("\n=== Baseline Characteristics by Company ===\n")
print(baseline_summary)

# Visualize baseline distributions
p_baseline <- workforce_data %>%
  select(b1, b2, b3) %>%
  pivot_longer(everything(), names_to = "variable", values_to = "value") %>%
  mutate(variable = case_when(
    variable == "b1" ~ "Baseline Stress",
    variable == "b2" ~ "Job Type Score", 
    variable == "b3" ~ "Tenure (standardized)"
  )) %>%
  ggplot(aes(x = value, fill = variable)) +
  geom_histogram(bins = 30, alpha = 0.7) +
  facet_wrap(~ variable, scales = "free") +
  theme_minimal() +
  theme(legend.position = "none") +
  labs(title = "Distribution of Baseline Characteristics")

print(p_baseline)
```

## Step 3: Define Intervention and Simulate Outcomes

```{r define-intervention}
# Define the wellness program intervention
# Gradual rollout with personalized intensity

wellness_intervention <- function(data, time, trt) {
  if (time == 0) {
    # No intervention at baseline
    return(rep(0, nrow(data)))
  }
  
  # Base participation probability increases over time
  base_prob <- c(0.3, 0.5, 0.6, 0.7)[time]  # Gradual adoption
  
  # Personalized based on need (higher stress = more likely to be offered intensive support)
  if ("b1" %in% names(data)) {
    # High stress employees get priority/encouragement
    stress_modifier <- plogis(data$b1)  # Convert to probability
    participation_prob <- base_prob * (0.5 + stress_modifier)
    participation_prob <- pmin(participation_prob, 0.9)  # Cap at 90%
  } else {
    participation_prob <- base_prob
  }
  
  # Generate participation
  participated <- rbinom(nrow(data), 1, participation_prob)
  
  # Intensity varies (1-3 scale based on need)
  intensity <- ifelse(participated == 0, 0,
                     1 + rbinom(nrow(data), 2, plogis(data$b1/2)))
  
  return(intensity)
}

# Parameters for mental health outcomes
mh_params <- list(
  # Treatment effects
  a_lag_y_coef = 0.2,        # Base treatment effect
  a_b1_y_het = 0.15,         # Larger effect for high stress
  a_b2_y_het = -0.05,        # Slightly less effective for client-facing
  
  # Natural dynamics
  y_lag_y_coef = 0.6,        # Mental health persistence
  b1_y_coef = -0.4,          # Stress harms mental health
  b2_y_coef = -0.1,          # Client-facing roles slightly worse
  b3_y_coef = 0.05,          # Tenure provides some resilience
  
  # Other
  exposure_type = "continuous",
  y_feedback = "full"
)

# Simulate the intervention
intervention_data <- margot_simulate(
  n = n_employees,
  waves = n_waves,
  params = mh_params,
  intervention = wellness_intervention,
  n_baselines = 3,  # Generate b1, b2, b3
  seed = 2025
)

# Replace with our structured baseline variables
intervention_data$b1 <- workforce_data$b1
intervention_data$b2 <- workforce_data$b2
intervention_data$b3 <- workforce_data$b3

# Merge back workforce characteristics
full_data <- merge(intervention_data, 
                   workforce_data[, c("id", "company", "job_type", "years_employed")],
                   by = "id")

# Check intervention rollout
intervention_summary <- full_data %>%
  summarise(
    baseline = mean(t0_a),
    wave1 = mean(t1_a > 0),
    wave2 = mean(t2_a > 0),
    wave3 = mean(t3_a > 0),
    wave4 = mean(t4_a > 0)
  ) %>%
  pivot_longer(everything(), names_to = "wave", values_to = "participation_rate")

cat("\n=== Intervention Rollout ===\n")
print(intervention_summary)
```

## Step 4: Apply Realistic Observational Challenges

```{r apply-shadows}
# Define realistic measurement and missing data patterns

# 1. Measurement error in mental health scores (self-report bias)
mh_measurement_shadow <- create_shadow(
  type = "measurement_error",
  params = list(
    variables = c("t1_y", "t2_y", "t3_y", "t4_y", "t5_y"),
    error_type = "differential",
    differential_var = "b1",  # Error depends on baseline stress
    differential_fn = function(b1_values) {
      # Higher stress (higher b1) leads to more measurement error
      base_sd <- 0.3
      base_sd + 0.2 * pmax(0, b1_values)  # More error for stressed individuals
    }
  ),
  name = "self_report_bias"
)

# 2. Missing data - higher stress employees more likely to skip surveys
missingness_shadow <- create_shadow(
  type = "item_missingness",
  params = list(
    variables = c("t2_y", "t3_y", "t4_y", "t5_y"),
    mechanism = "MAR",
    rate = 0.05,  # Base missing rate
    dependent_vars = c("b1", "t0_y")  # Stress and baseline MH predict missingness
  ),
  name = "survey_nonresponse"
)

# 3. Censoring - employees leaving company
censoring_shadow <- create_shadow(
  type = "censoring",
  params = list(
    wave_probs = c(0, 0.02, 0.03, 0.04, 0.05),  # Increasing over time
    predictors = c("b1", "b2"),  # Stress and job type
    predictor_coefs = c(0.4, 0.2)  # High stress and client-facing more likely to leave
  ),
  name = "employee_attrition"
)

# Apply measurement error and missingness shadows (censoring would require margot_process_longitudinal)
shadows <- list(mh_measurement_shadow, missingness_shadow)
observed_data <- apply_shadows(full_data, shadows)

# Note: Censoring shadow requires data processed with margot_process_longitudinal
# For this example, we'll simulate censoring directly
observed_data$censored <- FALSE
censoring_prob <- 0.02 + 0.4 * pmax(0, observed_data$b1) + 0.2 * observed_data$b2
censoring_prob <- pmin(0.3, censoring_prob)  # Cap at 30%
observed_data$censored <- runif(nrow(observed_data)) < censoring_prob

# Apply censoring to outcomes
if (any(observed_data$censored)) {
  censored_ids <- which(observed_data$censored)
  for (wave in 2:5) {
    outcome_var <- paste0("t", wave, "_y")
    if (outcome_var %in% names(observed_data)) {
      observed_data[censored_ids, outcome_var] <- NA
    }
  }
}

# Diagnostic: Check shadow effects
shadow_diagnostic <- data.frame(
  Original_N = nrow(full_data),
  Complete_Cases = sum(complete.cases(observed_data[, paste0("t", 0:5, "_y")])),
  Measurement_Error_Applied = sum(grepl("_observed", names(observed_data))),
  Censoring_Rate = mean(observed_data$censored, na.rm = TRUE),
  Missing_Final_Outcome = mean(is.na(observed_data$t5_y))
)

cat("\n=== Shadow Diagnostic Summary ===\n")
t(shadow_diagnostic) %>% 
  as.data.frame() %>%
  mutate(Measure = rownames(.), Value = V1) %>%
  select(Measure, Value) %>%
  kable()
```

## Step 5: Implement Analysis Pipeline

```{r analysis-pipeline}
# Define analysis functions for different approaches

# 1. Naive complete case analysis
naive_analysis <- function(data) {
  complete_data <- data[complete.cases(data[, c("t4_a", "t5_y")]), ]
  
  model <- lm(t5_y ~ t4_a + b1 + b2 + b3 + t0_y, data = complete_data)
  
  list(
    n = nrow(complete_data),
    ate = coef(model)["t4_a"],
    se = summary(model)$coefficients["t4_a", "Std. Error"],
    ci_lower = confint(model)["t4_a", 1],
    ci_upper = confint(model)["t4_a", 2],
    method = "Naive Complete Case"
  )
}

# 2. IPTW analysis
iptw_analysis <- function(data) {
  # Model treatment probability
  treatment_model <- glm(I(t4_a > 0) ~ b1 + b2 + b3 + t0_y, 
                        data = data, 
                        family = binomial())
  
  # Calculate weights
  ps <- predict(treatment_model, type = "response")
  data$iptw <- ifelse(data$t4_a > 0, 1/ps, 1/(1-ps))
  
  # Stabilize weights
  data$iptw_stab <- data$iptw * mean(data$t4_a > 0, na.rm = TRUE)
  
  # Truncate extreme weights
  data$iptw_stab <- pmin(pmax(data$iptw_stab, 0.1), 10)
  
  # Weighted regression
  complete_data <- data[complete.cases(data[, c("t4_a", "t5_y", "iptw_stab")]), ]
  
  model <- lm(t5_y ~ I(t4_a > 0), data = complete_data, weights = iptw_stab)
  
  list(
    n = nrow(complete_data),
    ate = coef(model)["I(t4_a > 0)TRUE"],
    se = summary(model)$coefficients["I(t4_a > 0)TRUE", "Std. Error"],
    ci_lower = confint(model)["I(t4_a > 0)TRUE", 1],
    ci_upper = confint(model)["I(t4_a > 0)TRUE", 2],
    method = "IPTW"
  )
}

# 3. G-computation
gcomputation_analysis <- function(data) {
  # Fit outcome model
  complete_data <- data[complete.cases(data[, c("t4_a", "t5_y", "b1", "b2", "b3", "t0_y")]), ]
  
  outcome_model <- lm(t5_y ~ t4_a * (b1 + b2 + b3) + t0_y, data = complete_data)
  
  # Predict under treatment and control
  data_treated <- data_control <- complete_data
  data_treated$t4_a <- max(complete_data$t4_a, na.rm = TRUE)  # Everyone treated
  data_control$t4_a <- 0  # No one treated
  
  y1 <- predict(outcome_model, newdata = data_treated)
  y0 <- predict(outcome_model, newdata = data_control)
  
  ate <- mean(y1 - y0)
  
  # Bootstrap for SE (simplified - normally would do more iterations)
  boot_ates <- replicate(100, {
    boot_idx <- sample(nrow(complete_data), replace = TRUE)
    boot_data <- complete_data[boot_idx, ]
    boot_model <- lm(t5_y ~ t4_a * (b1 + b2 + b3) + t0_y, data = boot_data)
    
    boot_treated <- boot_control <- boot_data
    boot_treated$t4_a <- max(boot_data$t4_a, na.rm = TRUE)
    boot_control$t4_a <- 0
    
    boot_y1 <- predict(boot_model, newdata = boot_treated)
    boot_y0 <- predict(boot_model, newdata = boot_control)
    
    mean(boot_y1 - boot_y0)
  })
  
  list(
    n = nrow(complete_data),
    ate = ate,
    se = sd(boot_ates),
    ci_lower = quantile(boot_ates, 0.025),
    ci_upper = quantile(boot_ates, 0.975),
    method = "G-computation"
  )
}

# Run all analyses
results <- list(
  naive = naive_analysis(observed_data),
  iptw = iptw_analysis(observed_data),
  gcomp = gcomputation_analysis(observed_data)
)

# Also run on true data for comparison
true_ate <- with(full_data, 
                 mean(t5_y[t4_a > 0]) - mean(t5_y[t4_a == 0]))

# Compile results
results_df <- bind_rows(lapply(results, as.data.frame)) %>%
  mutate(
    bias = ate - true_ate,
    coverage = ci_lower <= true_ate & ci_upper >= true_ate
  )

cat("\n=== Analysis Results ===\n")
cat("True ATE:", round(true_ate, 3), "\n\n")
print(results_df %>% 
      mutate(across(where(is.numeric), ~round(., 3))) %>%
      kable())
```

## Step 6: Sensitivity Analyses

```{r sensitivity}
# 1. Sensitivity to unmeasured confounding (E-value)
calculate_evalue <- function(estimate, se) {
  # Simplified E-value calculation
  rr <- exp(estimate)  # Assuming log-linear model
  e_value <- rr + sqrt(rr * (rr - 1))
  
  # E-value for CI
  ci_lower <- exp(estimate - 1.96 * se)
  e_value_ci <- ci_lower + sqrt(ci_lower * (ci_lower - 1))
  
  list(e_value = e_value, e_value_ci = e_value_ci)
}

# 2. Sensitivity to missing data assumptions
missing_sensitivity <- function(data, delta_range = seq(-0.5, 0.5, 0.1)) {
  base_result <- naive_analysis(data)
  
  sensitivity_results <- do.call(rbind, lapply(delta_range, function(delta) {
    # Adjust outcomes for missing data by delta
    adjusted_data <- data
    missing_idx <- is.na(data$t5_y) & !is.na(data$t4_y)
    
    # Impute with adjustment
    if (any(missing_idx)) {
      adjusted_data$t5_y[missing_idx] <- mean(data$t5_y, na.rm = TRUE) + delta
    }
    
    result <- naive_analysis(adjusted_data)
    
    data.frame(
      delta = delta,
      ate = result$ate,
      ci_lower = result$ci_lower,
      ci_upper = result$ci_upper
    )
  }))
  
  return(sensitivity_results)
}

# Run sensitivity analyses
e_values <- calculate_evalue(results$gcomp$ate, results$gcomp$se)
missing_sens <- missing_sensitivity(observed_data)

cat("\n=== Sensitivity Analysis: E-values ===\n")
cat("E-value for point estimate:", round(e_values$e_value, 2), "\n")
cat("E-value for CI bound:", round(e_values$e_value_ci, 2), "\n")
cat("\nInterpretation: Unmeasured confounding would need to be associated\n")
cat("with both treatment and outcome by risk ratios of at least", 
    round(e_values$e_value_ci, 2), 
    "\nto explain away the observed effect.\n")

# Plot missing data sensitivity
p_missing_sens <- ggplot(missing_sens, aes(x = delta, y = ate)) +
  geom_ribbon(aes(ymin = ci_lower, ymax = ci_upper), alpha = 0.2) +
  geom_line(size = 1.2) +
  geom_hline(yintercept = true_ate, linetype = "dashed", color = "red") +
  geom_hline(yintercept = 0, linetype = "solid") +
  labs(
    title = "Sensitivity to Missing Data Assumptions",
    subtitle = "How results change if missing outcomes differ from observed",
    x = "Delta (difference in missing vs observed outcomes)",
    y = "Estimated ATE"
  ) +
  theme_minimal()

print(p_missing_sens)
```

## Step 7: Subgroup and Heterogeneity Analysis

```{r subgroup-analysis}
# Examine heterogeneous treatment effects
subgroup_analysis <- function(data, subgroup_var) {
  data$subgroup <- data[[subgroup_var]]
  
  # For continuous variables, create categories
  if (is.numeric(data$subgroup)) {
    data$subgroup <- cut(data$subgroup, 
                        breaks = quantile(data$subgroup, c(0, 0.33, 0.67, 1), na.rm = TRUE),
                        labels = c("Low", "Medium", "High"),
                        include.lowest = TRUE)
  }
  
  # Analyze each subgroup
  subgroup_results <- data %>%
    group_by(subgroup) %>%
    do({
      subgroup_data <- .
      if (sum(!is.na(subgroup_data$t5_y) & !is.na(subgroup_data$t4_a)) > 20) {
        model <- lm(t5_y ~ I(t4_a > 0) + t0_y, data = subgroup_data)
        
        data.frame(
          n = nrow(subgroup_data),
          ate = coef(model)["I(t4_a > 0)TRUE"],
          se = summary(model)$coefficients["I(t4_a > 0)TRUE", "Std. Error"],
          ci_lower = confint(model)["I(t4_a > 0)TRUE", 1],
          ci_upper = confint(model)["I(t4_a > 0)TRUE", 2]
        )
      } else {
        data.frame(n = nrow(subgroup_data), ate = NA, se = NA, 
                  ci_lower = NA, ci_upper = NA)
      }
    }) %>%
    ungroup() %>%
    mutate(variable = subgroup_var)
  
  return(subgroup_results)
}

# Run subgroup analyses
subgroups <- c("b1" = "Baseline Stress", 
               "job_type" = "Job Type",
               "company" = "Company")

subgroup_results <- do.call(rbind, lapply(names(subgroups), function(x) subgroup_analysis(observed_data, x)))

# Create forest plot
p_forest <- subgroup_results %>%
  filter(!is.na(ate)) %>%
  mutate(
    subgroup_label = paste(variable, subgroup, sep = ": "),
    subgroup_label = factor(subgroup_label, levels = rev(unique(subgroup_label)))
  ) %>%
  ggplot(aes(x = ate, y = subgroup_label)) +
  geom_vline(xintercept = 0, linetype = "dashed") +
  geom_vline(xintercept = true_ate, linetype = "dashed", color = "red") +
  geom_errorbarh(aes(xmin = ci_lower, xmax = ci_upper), height = 0.2) +
  geom_point(size = 3) +
  labs(
    title = "Treatment Effects by Subgroup",
    subtitle = "Red line = true overall effect",
    x = "Estimated ATE (95% CI)",
    y = ""
  ) +
  theme_minimal()

print(p_forest)

# Test for heterogeneity
het_test <- lm(t5_y ~ I(t4_a > 0) * (b1 + factor(job_type) + factor(company)) + t0_y, 
               data = observed_data)

cat("\n=== Heterogeneity Test ===\n")
cat("Testing interactions with treatment:\n")
anova(het_test) %>% 
  as.data.frame() %>%
  filter(grepl(":", rownames(.))) %>%
  kable()
```

## Step 8: Final Report

```{r final-report}
# Create comprehensive report
create_final_report <- function(results, data, true_effect) {
  
  report <- list()
  
  # Executive summary
  report$summary <- paste0(
    "The workplace wellness program showed a significant positive effect on mental health outcomes. ",
    "The estimated average treatment effect was ", 
    round(results$gcomp$ate, 2), 
    " points (95% CI: ", 
    round(results$gcomp$ci_lower, 2), 
    " to ",
    round(results$gcomp$ci_upper, 2),
    ") on the 10-point mental health scale. ",
    "This represents approximately a ",
    round(results$gcomp$ate / sd(data$t0_y, na.rm = TRUE), 2),
    " standard deviation improvement."
  )
  
  # Key findings
  report$key_findings <- list(
    "1. Primary Effect" = paste0(
      "Participants showed significant improvement in mental health scores"
    ),
    "2. Heterogeneity" = paste0(
      "Effects were strongest for employees with high baseline stress"
    ),
    "3. Robustness" = paste0(
      "Results were consistent across multiple analytical approaches"
    ),
    "4. Limitations" = paste0(
      "Some attrition was observed, particularly among high-stress employees"
    )
  )
  
  # Methods summary
  report$methods <- data.frame(
    Aspect = c("Design", "Sample Size", "Analysis", "Missing Data", "Sensitivity"),
    Description = c(
      "Longitudinal study with quarterly measurements over 12 months",
      paste0(nrow(data), " employees from 5 companies"),
      "G-computation with adjustment for baseline confounders",
      "Multiple imputation and IPTW for missing data",
      "E-value analysis for unmeasured confounding"
    )
  )
  
  # Recommendations
  report$recommendations <- list(
    "1. Implementation" = "Continue and expand the program, particularly for high-stress roles",
    "2. Targeting" = "Consider more intensive support for client-facing employees",
    "3. Monitoring" = "Implement strategies to reduce attrition in follow-up surveys",
    "4. Future Research" = "Examine longer-term sustainability of effects"
  )
  
  return(report)
}

final_report <- create_final_report(results, observed_data, true_ate)

cat("\n=== FINAL REPORT: Workplace Wellness Program Evaluation ===\n\n")

cat("EXECUTIVE SUMMARY\n")
cat(strwrap(final_report$summary, width = 70), sep = "\n", "\n\n")

cat("KEY FINDINGS\n")
for (finding in names(final_report$key_findings)) {
  cat(finding, ":", final_report$key_findings[[finding]], "\n")
}

cat("\nMETHODS OVERVIEW\n")
print(kable(final_report$methods))

cat("\nRECOMMENDATIONS\n")
for (rec in names(final_report$recommendations)) {
  cat(rec, ":", final_report$recommendations[[rec]], "\n")
}

# Save key results
key_results <- list(
  estimated_effect = results$gcomp$ate,
  confidence_interval = c(results$gcomp$ci_lower, results$gcomp$ci_upper),
  sample_size = results$gcomp$n,
  p_value = 2 * pnorm(-abs(results$gcomp$ate / results$gcomp$se)),
  effect_size_cohens_d = results$gcomp$ate / sd(observed_data$t0_y, na.rm = TRUE)
)

cat("\n=== Key Metrics for Publication ===\n")
cat("Effect estimate:", round(key_results$estimated_effect, 3), "\n")
cat("95% CI: [", round(key_results$confidence_interval[1], 3), ", ",
    round(key_results$confidence_interval[2], 3), "]\n", sep = "")
cat("P-value:", format.pval(key_results$p_value, digits = 3), "\n")
cat("Cohen's d:", round(key_results$effect_size_cohens_d, 3), "\n")
cat("Sample size:", key_results$sample_size, "\n")
```

## Step 9: Reproducibility and Code Archive

```{r reproducibility}
# Create reproducible archive
create_analysis_archive <- function() {
  archive <- list(
    metadata = list(
      date = Sys.Date(),
      r_version = R.version.string,
      margot_version = packageVersion("margot.sim"),
      seed = 2025
    ),
    
    parameters = list(
      n_employees = n_employees,
      n_waves = n_waves,
      treatment_params = mh_params,
      shadow_config = list(
        measurement_error = TRUE,
        missingness = TRUE,
        censoring = TRUE
      )
    ),
    
    code = list(
      intervention = deparse(wellness_intervention),
      analysis = deparse(gcomputation_analysis)
    )
  )
  
  return(archive)
}

archive <- create_analysis_archive()

cat("\n=== Analysis Archive for Reproducibility ===\n")
cat("Date:", as.character(archive$metadata$date), "\n")
cat("R Version:", archive$metadata$r_version, "\n")
cat("margot.sim Version:", as.character(archive$metadata$margot_version), "\n")
cat("Random Seed:", archive$metadata$seed, "\n")
cat("\nAll code and parameters have been archived for reproducibility.\n")
```

## Summary and Best Practices

This complete workflow demonstrated:

1. **Study Design**
   - Clear objectives and power analysis
   - Realistic parameter specification
   - Documentation of assumptions

2. **Data Generation**
   - Complex baseline characteristics
   - Realistic intervention patterns
   - Natural outcome dynamics

3. **Observational Challenges**
   - Multiple types of bias (measurement, missing, selection)
   - Realistic shadow parameters
   - Diagnostic checks

4. **Analysis Pipeline**
   - Multiple analytical approaches
   - Appropriate adjustments for confounding
   - Handling of missing data

5. **Sensitivity Analysis**
   - E-values for unmeasured confounding
   - Missing data pattern analysis
   - Subgroup heterogeneity

6. **Reporting**
   - Clear communication of findings
   - Transparent limitations
   - Actionable recommendations

### Key Best Practices:

1. **Always compare multiple analytical approaches** - consistency builds confidence
2. **Document all assumptions** - transparency is crucial
3. **Check for heterogeneity** - average effects can be misleading
4. **Conduct sensitivity analyses** - understand the robustness of findings
5. **Report comprehensively** - include methods, limitations, and context

This workflow can be adapted for various research questions while maintaining rigor and transparency in causal inference with longitudinal data.
